The effect of symbols in government communications on citizens' trust, and its interaction with substantive information

### 2018-07-01 paper2_information and symbols_01.Rmd
Add files via upload

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements can increase citizens' trust via two mechanisms: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by leading citizens to dedicate less attention to the substantive information in the communication, and thus causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust. I examine these expectations via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their undervaluation of public-service performance. One of the main remedies which has been proposed for this problem is greater investment in direct government communications with citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@deacon_2001; @eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

The public administration research on the effect of government public communications on citizens attitudes has focused mostly on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on organizations' performance and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

@@

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a normative view of democratic theory, the potential effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Mainly, we should be concerned about the possibility that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust.

@@

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to think about its content. I discuss the theoretical and normative implications of these findings.

@@

Symbols are fundamental for almost any form of human interaction, as acknowledged by early scholars of social science [@blumer_1986]. A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. My focus is on those symbols that are strategically used by government organizations in their public communications, and are designed to distinguish organizations and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can be also found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

By comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the name of US federal agencies not only represents information about their activities, but may also carry specific symbolic meaning and affect that can shape citizens' support for government policy (depending on citizens' political party identification). These two studies provide examples for the role of symbols in shaping citizens' attitudes. Still, they did not effectively disentangle the effects of the symbolic and informational aspects of the studied public communications.

@@

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared between citizens who were personally exposed to prolonged power outages, and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the assessed government quality.

@@

ELM enables to disentangle the distinct effects of different elements in communications. Some elements are processed differently by people at different locations along the elaboration continuum. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as arguments, and systematically scrutinize it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all. For example, the physical attractiveness of the source normally does not represent any substantive argument, yet it can serve as a simple cue. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

ELM further postulates that the placement of people along the elaboration continuum in any given situation is determined by their *motivation* and *ability* to think about the message. Accordingly, the literature proposes a variety of such motivational and ability factors. Among others, studies have demonstrated that the more people are distracted, the more difficult it is for them to scrutinize the message arguments, a mechanism that has been referred to as "elaboration disruption" [@petty_1976; @petty_1986]. Another prominent factor that influences people's inclination to elaborate on the message is their affective state, or mood. Positive mood (compared with neutral or negative moods) reduces people's cognitive ability and attenuates their motivation to think, due to their desire to maintain their positive mood and avoid message processing that would depress it [@worth_1987; @mackie_1989; @bless_1990; @schwarz_1991; @petty_2015; but see: @wegener_1995].[^footnote_elm_factors] Affecting the amount of thinking can cause attitude change in interaction with elements in the communication that can function as arguments or peripheral cues. Decreasing the amount of thinking weakens the effect the substantive arguments and enhances the effect of peripheral cues (and vice versa).

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]; The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuades people directly through the peripheral route, yet at the same time they can also as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust through the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).

Symbolic elements may decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and absorbing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message - government organizations and policies - is likely to yield negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood. The latter expectation entails an interaction between symbols and the substantive information in the communications. Symbolic elements are likely to increase citizens' trust insofar as they reduce citizens' inclination to scrutinize information that would have decreased their trust.

@@

>*H<sub>1</sub> - Symbolic elements in public organizations' communications are likely to increase citizens' trust in the organizations and in the policies presented in the communications.*

>*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that would have decreased citizens' trust.*

>*H<sub>3</sub> - Symbolic elements are likely to decrease the amount of thinking about the content of the communication.*

>*H<sub>4</sub> - The abovementioned conditioned effect of symbolic elements on citizens' trust is likely to be mediated by the decrease in the amount of thinking (a moderated mediation hypothesis).*

@@

Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans. Nevertheless, one can make the argument that neutralizing the effect of "positive" information with symbols may not actually yield significant differences, since citizens are more likely to "punish" bureaucracies for publishing negative information than to "praise" them for publishing positive information [@james_2011; @nielsen_2017; @james_2017b]. Congruently, the effect of positive information (versus a medium level) is likely to be relatively small, and hence neutralizing this effect may not make much difference. Given the focus of this study on the interaction between symbols and ill-conceived policies, the research design does not directly addresses this latter interaction.

@@

In addition to the above-mentioned hypotheses, the experiment is also designed to test an additional set of hypotheses regarding the interaction between symbols and the perceived personal relevance of communications. The selection of policy plans and the construction of the population sample were designed to account for variation in that variable. In this paper, I focus on those methodological and empirical parts of the study that are consequential for the examination of the interaction between symbols and information.[^footnote_relevance_paper]

@@

Empirically, I focused on the case of the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, both at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM makes a good case for this study since has invested greatly over the past years in public communications. In these various communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (image a. in figure 1). This logo consists of pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element in which I focused is the green color, which is strongly associated with taking care of the environment. The EPM's logo and the green color can be seen in almost all of the visual communications of the EPM, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, that previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (image b.) was launched in 2010, and included a series of advertisements in which Friedman was presented with a green grass head, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (image c.) was launched in 2017, and was focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

@@

The survey was presented to the participants as a study about citizens' attitudes regarding environmental policy issues. The participants were first asked pre-manipulation questions about their political ideology and their perceptions about the pubic sector.[^footnote_relevance_manipulation] Thereafter, they were told that they will be presented with two government policy plans regarding specific environmental issues, obtained from EPM's official publication of its annual work plan for the year 2018. The policy plans regarded to reducing air-pollution in the Bay of Haifa (an area in Israel) and reducing domestic waste and increasing recycling. The two policy plans were presented to each subject in a random order. Each policy plan included the title of the policy goal (i.e. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

@@

Per each policy, the subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). The policy activities which were mentioned in the strong policy plans condition were taken from the real EPM work plan and from other reports and official sources of EPM. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures which are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were increasing the supervision of polluting factories, and raising citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is displayed in **Appendix A**. For each subject, one of the two policies was strong and the other was weak. For example, a subject who first received a strong air-pollution policy, later received the weak recycling policy, and vice versa.

In addition to information manipulation, the participants are randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment (real symbols), control no symbols and control fake symbols. In the symbol treatment condition, the two policies (again, one weak and one strong) are displayed with the above-mentioned three symbolic elements: the titles and the footer are colored with green, EPM logo is displayed at the top of the page, and the celebrities' images from the campaigns are attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags is attached to the recycling policy; Friedman's image from the "starting to think green" campaign was attached to the air-pollution policy. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities images. Finally, in the control fake-symbols condition, the green color is replaced with blue, the EPM logo is replaced by a fake logo and the images of celebrities from the campaigns are replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition is to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols to that they will resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which can be attached to the government ministry and environmental issues.[^footnote_symbols_manipulation_checks] In that sense, it can be regarded as a placebo. The symbols conditions are displayed in **Appendix B**. Further details about the images in the real and fake symbols conditions are found in the supplementary appendix. That each subject received two policies which were presented under the same symbol condition, but under different information conditions (i.e. one strong and one weak), enables me later to analyze the interaction between these two manipulations both via between- and within-subjects.

@@

I now turn to analyzing the experimental results. I begin by examining my first and second hypotheses, that regard the effect of symbols on trust and its interaction with information. Thereafter, I will examine hypotheses 3 and 4, that regard the mediating role on elaboration. For each section, I conducted both *between-subjects* and *within-subjects* analyses.

@@

In **Table 2**, I test the effects of the real symbols on the elaboration measures via regression analyses. I regressed each measure on the symbols manipulation (Models 2.1, 2.3), and than added the information manipulation (Models 2.2, 2.4). The first two models, which refer to reaction time, are random intercept models which are similar to the statistical models in table 1. As for the memory score analyses, given that each subject was asked only about one policy (the second in order), I used an OLS regression.

>>>>

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements can increase citizens' trust via two mechanisms: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by leading citizens to dedicate less attention to the substantive information in the communication, and thus causing them to disregard the content of logically unpersuasive information that would have otherwise decreased their trust. I examine these expectations via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was stronger among those who were exposed to the logically unpersuasive policy plans, yet distracted by the symbols and thus less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their underestimation of public-service performance. One of the main remedies which has been proposed for this problem is greater investment in direct government communications with citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@deacon_2001; @eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant public administration research on the effect of government public communications on citizens attitudes has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on organizations' performance and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

@@

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a democratic theory standpoint, the effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Mainly, we should be concerned about the possibility that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard the content of logically unpersuasive information that would have otherwise decreased their trust.

@@

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, yet less inclined to think about the policy content due to distraction by the symbols. I discuss the theoretical and normative implications of these findings.

@@

Symbols are fundamental for almost any form of human interaction, as acknowledged by early scholars of social science [@blumer_1986]. A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. My focus is on those symbols that are strategically used by government organizations in their public communications, are designed to distinguish them and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can be also found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

By comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the names of US federal agencies not only represent information about their activities, but may also carry specific symbolic meaning and affect that can shape citizens' support for government policies (depending on citizens' political party identification). These two studies provide examples for the role of symbols in shaping citizens' attitudes. Still, they did not effectively disentangle the effects of the symbolic and informational aspects of the studied public communications.

@@

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared citizens who were personally exposed to prolonged power outages and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the assessed government quality.

@@

ELM enables to disentangle the distinct effects of different elements in communications. Some elements are processed differently by people at different locations along the elaboration continuum. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as argumentation, and systematically scrutinize it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all. For example, the physical attractiveness of the source normally does not represent any substantive argument, yet it can serve as a simple cue. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

ELM further postulates that the placement of people along the elaboration continuum in any given situation is determined by their *motivation* and *ability* to think about the message. Accordingly, the literature proposes a variety of such motivational and ability factors. Among others, studies have demonstrated that the more people are distracted, the more difficult it is for them to scrutinize the message arguments, a mechanism that has been referred to as "elaboration disruption" [@petty_1976; @petty_1986]. Another prominent factor that influences people's inclination to elaborate on the message is their affective state, or mood. Positive mood (compared with neutral or negative moods) reduces people's cognitive ability and attenuates their motivation to think, due to their desire to maintain their positive mood and avoid message processing that would depress it [@worth_1987; @mackie_1989; @bless_1990; @schwarz_1991; @petty_2015; but see: @wegener_1995].[^footnote_elm_factors] Affecting the amount of thinking can cause attitude change in interaction with elements in the communication that can function as arguments or peripheral cues. Decreasing the amount of thinking weakens the effect of substantive arguments and enhances the effect of peripheral cues (and vice versa).

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]. The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuades people directly through the peripheral route, yet at the same time they can also act as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust through the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).[^footnote_interaction_strong_information]

Symbolic elements may decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and absorbing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message - government organizations and policies - is likely to yield negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood. The latter expectation entails an interaction between symbols and the substantive information in the communications.

[^footnote_interaction_strong_information]: Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans. Given the focus of this study on the interaction between symbols and ill-conceived policies, the research design does not directly addresses this latter interaction.

@@

*H<sub>1</sub> - Symbolic elements in public organizations' communications are likely to increase citizens' trust in the organizations and in the policies presented in the communications.*

*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that would have decreased citizens' trust.*

*H<sub>3</sub> - Symbolic elements are likely to decrease the amount of thinking about the content of the communication.*

*H<sub>4</sub> - The abovementioned moderated effect of symbolic elements on citizens' trust H<sub>2</sub> is likely to be mediated by their decreasing effect on citizens’ amount of thinking H<sub>3</sub> (a moderated mediation hypothesis).*

@@

@@

In addition to the above-mentioned hypotheses, the experiment was also designed to test a different set of hypotheses regarding the interaction between symbols and the perceived personal relevance of communications. The selection of policy plans and the construction of the population sample were designed to account for variation in that variable. In this paper, however, I focus on those methodological and empirical parts of the study that are consequential for the examination of the interaction between symbols and information.[^footnote_relevance_paper]

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM makes a good case for this study due to its significant investment, over the past years, in public communications. In these various communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (image a. in figure 2). This logo consists a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element on which I focused is the green color, which is strongly associated with taking care of the environment. The EPM's logo and the green color can be seen in almost all of the visual communications of the EPM, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, that previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (image b.) was launched in 2010, and included a series of advertisements in which Friedman was presented with a green grass head, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (image c.) was launched in 2017, and was focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

@@

The survey was presented to the participants as a study about citizens' attitudes regarding environmental policy issues. The participants were first asked to respond to pre-manipulation questions about their political ideology and their perceptions of the pubic sector.[^footnote_relevance_manipulation] Thereafter, they were told that they will be presented with two government policy plans regarding specific environmental issues, obtained from EPM's official publication of its annual work plan for the year 2018. The policy plans regarded reducing air-pollution in the Bay of Haifa (an area in the north of Israel) and reducing domestic waste and increasing recycling. The two policy plans were presented to each subject in a random order. Each policy plan included the title of the policy goal (i.e. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

@@

Per each policy, subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). Each subject, moreover, was exposed to one strong and one weak policy communication. For example, a subject who first received a strong air-pollution policy, later received the weak recycling policy, and vice versa. The actions which were mentioned in the strong policy plans condition were taken from the real EPM work plan and from other reports and official sources of EPM. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures that are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were decreasing the supervision of polluting factories, and raising citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is displayed in **Appendix A**.

In addition to information manipulation, the participants were randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment ("real symbols"), control-"no symbols" and control-"fake symbols". In the symbol treatment condition, the two policies (again, one weak and one strong) are displayed with the above-mentioned three symbolic elements: the titles and the footer are colored with green, EPM logo is displayed at the top of the page, and the celebrities' images from the campaigns are attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags is attached to the recycling policy; Friedman's image from the "starting to think green" campaign was attached to the air-pollution policy. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities images. Finally, in the control fake-symbols condition, the green color is replaced with blue, the EPM logo is replaced by a fake logo and the images of celebrities from the campaigns are replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition is to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they will resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which can be attached to the government ministry and environmental issues.[^footnote_symbols_manipulation_checks] In that sense, it can be regarded as a placebo. The symbols conditions are displayed in **Appendix B**. Further details about the images in the real and fake symbols conditions are found in the supplementary appendix. That each subject received two policies which were presented under the same symbol condition, but under different information conditions (i.e. one strong and one weak), enables me later to analyze the interaction between these two manipulations both via between- and within-subjects.

@@

I now turn to analyzing the experimental results. I begin by examining my first and second hypotheses, that regard the effect of symbols on trust and its interaction with information. Thereafter, I will examine hypotheses 3 and 4, that regard the mediating role of elaboration. For each section, I conduct both *between-subjects* and *within-subjects* analyses.

@@

In **Table 2**, I test the effects of the real symbols on the elaboration measures via regression analyses. I regressed each elaboration measure on the symbols manipulation (Models 2.1, 2.3), and then added the information manipulation (Models 2.2, 2.4). The first two models, which refer to reaction time, are random intercept models which are similar to the statistical models in table 1. As for the memory score analyses, given that each subject was asked only about one policy (the second in order), I used an OLS regression.
